A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-01-30 analysis/analyses/4-method_characterisation/2_quality_control.Rmd
bye bye common.R, hello common in dynalyisis package

method_qc <- readRDS(derived_file("method_qc.rds"))
checks <- readRDS(derived_file("checks.rds"))

source("0_common.R")
```

@@

Orderin by category or application
```{r}
method_qc %>%
  group_by(method_id, category) %>%
  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
  ggplot(aes(method_id, category)) + geom_raster(aes(fill=score))

method_qc %>%
  gather(application, application_applicable, !!applications) %>%
  filter(application_applicable) %>%
  group_by(method_id, application) %>%
  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
  ggplot(aes(method_id, application)) + geom_raster(aes(fill=score))
```

@@

check_heatmap_plot <- method_qc %>%
  left_join(check_position, by="check_id") %>%
  mutate(method_id = factor(method_id)) %>%
  mutate(answer = ifelse(is.na(answer), 0, answer)) %>%
  ggplot() +
    geom_rect(aes(xmax=check_start, xmin=check_end, ymin=as.integer(method_id), ymax=as.integer(method_id) + 1, fill=category, alpha=answer)) +
    geom_vline(aes(xintercept=check_end), data=check_position, color="#444444", alpha=0.25) +
    geom_vline(aes(xintercept=question_end), data=question_position, color="#444444") +
    scale_x_reverse("Check", breaks = question_position$question_mid, labels = question_position$name, expand = c(0, 0)) +
    scale_y_continuous("Method", breaks = seq_along(levels(method_qc$method_id))+0.5, labels = levels(method_qc$method_id), expand = c(0, 0)) +
    scale_fill_manual(values=category_colors) +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

@@

method_qc_check_plot %>% saveRDS(figure_file("method_qc_check_plot.rds"))
```

@@

question_ordering_plot1 <- question_ordering_plotdata %>% 
  ggplot() +
    geom_bar(aes(question_id, n, fill=qc_score), stat="identity") +
    scale_fill_distiller(palette="Blues", direction = -1)
question_ordering_plot1

question_ordering_plotdata2 <- question_ordering_plotdata %>% 
  mutate(category = factor(category, levels=categories)) %>% 
  arrange(-as.numeric(category), question_ordering) %>% 
  mutate(question_id = factor(question_id, levels=as.character(unique(question_id))))

@@

question_ordering_plot2 <- question_ordering_plotdata2 %>% 
  arrange(qc_score) %>% 
  ggplot() +
    geom_bar(aes(question_id, n, fill=category_colors[category], alpha=qc_score), stat="identity") +
    geom_vline(aes(xintercept=start), data=question_ordering_plotdata2_category) +
    scale_fill_identity() +
    scale_alpha_continuous(range=c(1, 0.05), name = "Quality control score", guide = guide_legend(title.position = "top", title.hjust = 0.5,label.position = "bottom"))
question_ordering_plot2

category_legend <- question_ordering_plotdata2 %>% 
  group_by(category, question_id) %>% 
  summarise() %>% 
  ungroup() %>% 
  mutate(position = seq_len(n())) %>% 
  group_by(category) %>% 
  summarise(start = -min(position), end=-max(position)-1) %>% 
  ggplot() +
    geom_rect(aes(xmin=0, xmax=1, ymin=start, ymax=end), fill="white") +
    geom_text(aes(1, start + (end-start)/2, label=category_labels[category], color=category), hjust=1, size=5) +
    geom_segment(aes(x=1.1, xend=1.1, y=start-0.2, yend=end+0.2, color=category)) +
    scale_color_manual(values=category_colors) +
    scale_y_continuous(expand=c(0, 0)) +
    theme_void() +
    theme(legend.position="none")
category_legend

@@

method_qc_ordering_plot <- cowplot::plot_grid(category_legend, question_ordering_plot2, align = "h", rel_widths = c(0.2, 0.8), axis="tb")
method_qc_ordering_plot
saveRDS(method_qc_ordering_plot, figure_file("method_qc_ordering_plot.rds"))
```

>>>>

method_qc <- readRDS(derived_file("method_qc.rds"))
methods_evaluated <- readRDS(derived_file("methods_evaluated.rds"))
checks <- readRDS(derived_file("checks.rds"))
```

Comparing qc and citations overtime     
```{r, fig.width=16, fig.height=8}
g1 <- ggplot(methods_evaluated, aes(date, qc_score)) +
    geom_smooth(span=2) +
    geom_point() +
    ggrepel::geom_label_repel(aes(label = name)) +
    cowplot::theme_cowplot() +
    labs(x = "Time", y = "QC score", title = "Code quality score over time")
g1

g2 <- ggplot(methods_evaluated, aes(date, ncitations+1)) +
    geom_smooth(span=2) +
    geom_point() +
    ggrepel::geom_label_repel(aes(label = name)) +
    cowplot::theme_cowplot() +
    scale_y_log10() +
    labs(x = "Time", y = "Citations", title = "Citations over time")
g2

g3 <- ggplot(methods_evaluated, aes(ncitations+1, qc_score)) +
    geom_smooth(span=2) +
    geom_point() +
    ggrepel::geom_label_repel(aes(label = name)) +
    cowplot::theme_cowplot() +
    scale_x_log10() +
    labs(x = "Citations", y = "QC score", title = "QC score over # citations")
g3

cowplot::plot_grid(g1, g3, g2, ncol=2, nrow=2, labels = "auto") %>% 
  write_rds(figure_file("qc_over_time.rds"))
```

@@

Ordering by category or application
```{r}
method_qc %>%
  group_by(method_id, category) %>%
  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
  ggplot(aes(method_id, category)) + geom_raster(aes(fill=score))

method_qc %>%
  gather(application, application_applicable, !!qc_applications$application) %>%
  filter(application_applicable) %>%
  group_by(method_id, application) %>%
  summarise(score=sum(answer * score * weight)/sum(score * weight)) %>%
  ggplot(aes(method_id, application)) + geom_raster(aes(fill=score))
```

@@

check_heatmap_plot <- method_qc %>%
  left_join(check_position, by="check_id") %>%
  mutate(method_id = factor(method_id)) %>%
  mutate(answer = ifelse(is.na(answer), 0, answer)) %>%
  ggplot() +
    geom_rect(aes(xmax=check_start, xmin=check_end, ymin=as.integer(method_id), ymax=as.integer(method_id) + 1, fill=category, alpha=answer)) +
    geom_vline(aes(xintercept=check_end), data=check_position, color="#444444", alpha=0.25) +
    geom_vline(aes(xintercept=question_end), data=question_position, color="#444444") +
    scale_x_reverse("Check", breaks = question_position$question_mid, labels = question_position$name, expand = c(0, 0)) +
    scale_y_continuous("Method", breaks = seq_along(levels(method_qc$method_id))+0.5, labels = levels(method_qc$method_id), expand = c(0, 0)) +
    scale_fill_manual(values=set_names(qc_categories$color, qc_categories$category)) +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

@@

method_qc_check_plot %>% saveRDS(figure_file("method_qc_check.rds"))
```

@@

question_ordering_plotdata2 <- question_ordering_plotdata %>% 
  mutate(category = factor(category, levels=qc_categories$category)) %>% 
  arrange(-as.numeric(category), question_ordering) %>% 
  mutate(question_id = factor(question_id, levels=as.character(unique(question_id))))

@@

question_ordering_plot2 <- question_ordering_plotdata2 %>% 
  arrange(qc_score) %>% 
  ggplot() +
    geom_bar(aes(question_id, n, fill=set_names(qc_categories$color, qc_categories$category)[category], alpha=qc_score), stat="identity") +
    geom_vline(aes(xintercept=start), data=question_ordering_plotdata2_category) +
    scale_fill_identity() +
    scale_alpha_continuous(range=c(1, 0.05), name = "Quality control score", guide = guide_legend(title.position = "top", title.hjust = 0.5,label.position = "bottom"))
question_ordering_plot2

category_legend <- question_ordering_plotdata2 %>% 
  group_by(category, question_id) %>% 
  summarise() %>% 
  ungroup() %>% 
  mutate(position = seq_len(n())) %>% 
  group_by(category) %>% 
  summarise(start = -min(position), end=-max(position)-1) %>% 
  ggplot() +
    geom_rect(aes(xmin=0, xmax=1, ymin=start, ymax=end), fill="white") +
    geom_text(aes(1, start + (end-start)/2, label=label_long(category), color=category), hjust=1, size=5) +
    geom_segment(aes(x=1.1, xend=1.1, y=start-0.2, yend=end+0.2, color=category)) +
    scale_color_manual(values=set_names(qc_categories$color, qc_categories$category)) +
    scale_y_continuous(expand=c(0, 0)) +
    theme_void() +
    theme(legend.position="none")
category_legend

@@

method_qc_ordering_plot <- cowplot::plot_grid(category_legend, question_ordering_plot2, align = "h", rel_widths = c(0.2, 0.8), axis="tb")
method_qc_ordering_plot
saveRDS(method_qc_ordering_plot, figure_file("method_qc_ordering.rds"))
```

### 2018-01-30 analysis/paper/paper.Rmd
bye bye common.R, hello common in dynalyisis package

<!--- history of Ti, different methods --->
Dozens of trajectory inference methods have been developed over the last years, and more are being published almost every month (`r ref("sfig", "methods_over_time", "a")`). Initially, most trajectory inference was focused on linear trajectories, with one start and one end point. However, the prospect of branching trajectories was already described early [@trapnell_dynamics_2014], and about half of the current methods can handle one or more bifurcations in the trajectory (`r ref("sfig", "methods_over_time", "b")`). Some methods can already handle more complex, cyclical and branching behaviour [@welch_slicer:_2016; @wolf_graph_2017], and it is expected that in the future methods will be able to model even more complex behavior, such as multiple dynamic processes happening at parallel in a single-cell or the integration of datasets from different patients [@tanay_scaling_2017; @cannoodt_computational_2016].

@@

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface. We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).

`r ref("table", "methods", anchor=TRUE)` Characterisation and quality control of the evaluated trajectory inference methods
```{r}
read_rds(figure_file("methods_table.rds", "method_characteristics")) %>%
        knitr::kable()
```

At the level of trajectory structures, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`). However, only one method (SLICER) can currently handle more complex graph structures which can include loops (`r ref("sfig", "methods_over_time", "b")`).

@@

Most methods fulfilling most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`). Recent methods tend to have higher code quality sFIG. However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding studies pinpointing these as good practices [@wilson_best_2014].

@@

```{r}
add_sfig({
readRDS(figure_file("method_qc_check_plot.rds", experiment_id="method_characteristics"))
}, 
"method_qc_check", 
"Caption", 
10, 
13
)
```

```{r}
add_sfig({
readRDS(figure_file("method_components_plot.rds", experiment_id="method_characteristics"))
}, 
"method_components", 
"TI methods frequently share steps", 
15, 
7.5
)
```
### Evaluation overview
`r anchor("fig", "evaluation_overview")`
![](`r figure_file("ti_evaluation_overview_v1.svg", experiment_id="evaluation_overview")`)
`r ref("fig", "evaluation_overview")` Overview of the evaluation

@@

## Toy data generation

## Synthetic data generation

@@

pwalk(sfigs, function(ref_id, fig, caption, width, height) {
  subchunkify(ref_id, caption, width, height)
})
```

@@

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

>>>>

<!--- history of Ti, different methods --->

@@

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(methods$evaluated)` for evaluation, primarily based on their free availability and the presence of a programming interface `r ref("table", "methods")`. We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).

`r ref("table", "methods", anchor=TRUE)` Overview of current trajectory inference methods, and whether they were included in this current evaluation study
```{r}
read_rds(figure_file("methods_table.rds", "method_characteristics"))
```

At the level of trajectory structures, most recent methods can handle one or more splits (either bifurcations or multifurcations) in the trajectory (`r ref("fig", "method_characteristics", "a")`). However, only a handful of more recent methods can currently handle more complex graph structures which can include loops (`r ref("sfig", "methods_over_time", "b")`), one of which is included in the current evaluation (SLICER).

@@

Most methods fulfilling most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`). Recent methods tend to have higher code quality (`r ref("sfig", "qc_over_time")`). However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding studies pinpointing these as good practices [@wilson_best_2014].

@@

```{r}
add_sfig({
readRDS(figure_file("method_qc_check.rds", experiment_id="method_characteristics"))
}, 
"method_qc_check", 
"Caption", 
10, 
13
)
```

```{r}
add_sfig({
readRDS(figure_file("method_components.rds", experiment_id="method_characteristics"))
}, 
"method_components", 
"TI methods frequently share steps", 
15, 
7.5
)
```

```{r}
add_sfig({
readRDS(figure_file("qc_over_time.rds", experiment_id="method_characteristics"))
}, 
"qc_over_time", 
"Quality control scores and number of citations over time", 
10, 
10
)
```
### Evaluation overview
`r anchor("fig", "evaluation_overview")`
![](`r figure_file("ti_evaluation_overview_v1.svg", experiment_id="evaluation_overview")`)
`r ref("fig", "evaluation_overview")` Overview of the evaluation

@@

We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods `r ref("table", "methods", anchor=TRUE)`, based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018,@gitter_singlecellpseudotimeoverviewalgorithms_2018] and by searching in literature for "trajectory inference" and "pseudotemporal ordering". We believe this to be the most complete list to data, but are happy to include additional methods through e-mail or an issue on our github (https://github.com/dynverse/dynalysis). For each method we investigated whether it was 
## Synthetic data generation

@@

pwalk(sfigs %>% arrange(match(ref_id, refs$ref_id)), function(ref_id, fig, caption, width, height) {
  subchunkify(ref_id, caption, width, height)
})
```

@@

This report was generated on `r Sys.time()` using `r devtools:::platform_info()$version` and the following packages:
```{r colophon, cache = FALSE}
devtools::session_info()$packages %>% knitr::kable()
```
